---
title: 'COMPMUT Experiments 6: Presentation and analysis of dynamics data'
author: "jpjh"
date: "compiled Feb 2021"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())

library(tidyverse)
library(ggplot2)
library(knitr)

source("../functions/theme_pub.r")
theme_github <- function(x){
  theme_pub(base_size=12)
}

theme_set(theme_github())
```

## 'Medium-term' culture experiments with strains carrying different evolved pQBR57 variants in soil

#### Experimental design

The only parallel mutation identified in the long-term soil plasmid evolution experiment was in PQBR57_0059. Several different mutations to PQBR57_0059 were identified. To understand whether these mutants had a similar effect on plasmid dynamics, 'evolved' plasmids were conjugated into differentially-labelled ancestral strains of *P. fluorescens* SBW25. These plasmid-bearers were mixed in approximately equal numbers with plasmid-free ancestral strains, and the competitors were co-cultured in soil microcosms for five transfers, following the protocol in [Hall et al. 2016](dx.doi.org/10.1073/pnas.1600974113). The densities of each competitor were tracked by use of the *lacZ* marker present on the streptomycin-resistant-labelled strain, and plasmid carriage by each competitor was followed by replica plating onto mercury-containing media and identification of blue and white resistant colonies. To test whether mercury resistance correlated with plasmid carriage, selected colonies were tested by PCR at the end; no appreciable plasmid loss was detected in any population, consistent with the findings of [Hall et al. 2016](dx.doi.org/10.1073/pnas.1600974113).

#### Plotting of data

Load up data.

```{r}
d6 <- read.csv("../data/COMPMUT_exp_data_6.csv", header=TRUE, sep=",")
```

Calculate cfu/g soil for white and blue colonies. Remove timepoint 0, as this will not be included in the plot.

Note that I assume total release of bacteria from soil (some assumption) and estimate total liquid in the soil to be 11 ml (1 ml added from the previous transfer, 10 ml added when retrieving), hence a factor of (11/10) to get cfu/g rather than cfu/ml soil wash. 

For some of these experiments, the plasmid starts in the 'blue' streptomycin-resistant background, whereas in others it starts in the 'white' gentamicin-resistant background. This is labelled in the 'marker' column of the dataframe; 'Sm' indicates it starts in the blue background, 'Gm' in the white background.

Starting plasmid-bearers will be called 'donors', whereas starting plasmid-free will be called 'recipients'. Using this I will also identify 'segregants' (i.e. using the column 'hg_lost') — note that these are only segregants from the donor population, and may be reinfected with plasmid - and 'transconjugants' (using the column 'hg_kept'), with similar caveats.

For some samples, I also plated on antibiotics to detect small subpopulations. A preliminary examination of the data suggests it's appropriate to take the larger cfu/g count from populations that were plated out on different media.

```{r}
df6 <- d6 %>% filter(timepoint != 0) %>%
  mutate(cfu.g_white = (1000/spread) * (10^dilution) * count_white * (11/10),
         cfu.g_blue  = (1000/spread) * (10^dilution) * count_blue * (11/10),
         cfu.g_D = ifelse(marker == "Sm", cfu.g_blue, cfu.g_white),
         cfu.g_R = ifelse(marker == "Gm", cfu.g_blue, cfu.g_white),
         cfu.g_S = (1000/spread) * (10^dilution) * hg_lost * (11/10),
         cfu.g_T = (1000/spread) * (10^dilution) * hg_kept * (11/10)) %>%
  select(replicate, variant, marker, timepoint, media, cfu.g_D, cfu.g_R, cfu.g_S, cfu.g_T) %>%
  group_by(replicate, variant, timepoint, marker) %>%
  replace_na(list(cfu.g_D = 0, cfu.g_R = 0, cfu.g_S = 0, cfu.g_T = 0)) %>%
  summarise(cfu.g_D = max(cfu.g_D, na.rm=TRUE),
            cfu.g_R = max(cfu.g_R, na.rm=TRUE),
            cfu.g_S = max(cfu.g_S, na.rm=TRUE),
            cfu.g_T = max(cfu.g_T, na.rm=TRUE)) 
```

Get the data for the ancestral plasmid. This was previously published in [Hall et al. 2017](dx.doi.org/10.1038/s41559-017-0250-3).

```{r}
d_anc <- read.csv("../ext/S2_Hg_transfer_data.csv", header=TRUE, sep=",") %>%
  filter(mercury==0 & transfer > 0) %>%
  transmute(replicate = factor(rep, levels=letters[1:6],
                               labels = c("a","b","c","a","b","c")),
         timepoint = transfer,
         marker = donor_marker,
         cfu.g_D = donor_density,
         cfu.g_R = recipient_density,
         cfu.g_S = donor_density - donor_plasmid_density,
         cfu.g_T = recipient_plasmid_density,
         variant = ifelse(rep %in% c("d","e","f"),
                          "unevolved ", "unevolved"))
```

Join tables and calculate ratios.

Note that in the original tables, 'donor' includes segregants, and 'recipient' includes transconjugants. When calculating ratios need to ensure that this is accounted for.

```{r}
dl6 <- bind_rows(d_anc, df6) %>% pivot_longer(cols = contains("cfu"), names_to = "subpopulation", values_to = "cfu.g")

dr6 <- bind_rows(d_anc, df6) %>% 
  mutate(ratio_D = (cfu.g_D-cfu.g_S)/(cfu.g_D + cfu.g_R),
         ratio_R = (cfu.g_R-cfu.g_T)/(cfu.g_D + cfu.g_R),
         ratio_S = (cfu.g_S)/(cfu.g_D + cfu.g_R),
         ratio_T = (cfu.g_T)/(cfu.g_D + cfu.g_R)) %>%
  pivot_longer(cols = contains("ratio"), names_to = "subpopulation", values_to = "ratio") %>%
  select(-contains("cfu"))
```

Plot:

```{r}
dl6 %>% ggplot(aes(x=timepoint, y=cfu.g, colour=subpopulation)) + 
  geom_line() + facet_grid(replicate~variant)
```

Try a stacked area chart using the ratios.

```{r}
dr6 %>% ggplot(aes(x=timepoint, y=ratio, fill=subpopulation)) +
  geom_area(position="stack") + facet_grid(replicate~variant)
```

Tidy up the plot: sort out colours, order panels, etc.

Put the ancestral plasmids on the left, then the evolved plasmids with ancestral-like PQBR57_0059, then the evolved plasmids with mutated PQBR57_0059.

For the ratios, put plasmid-free compartments at the top and bottom, and plasmid-containing compartments in the middle.

```{r}
variants <- data.frame(v = c("unevolved","unevolved ","A01","A11","C01","C10","C11",
                             "A04","B01","B02","B03","B09","B10","C03"),
                       PQ5759 = c(rep("anc",7), rep("mut",7)))

dr6 <- dr6 %>% mutate(subpopulation = factor(subpopulation, levels=c("ratio_R","ratio_T","ratio_D","ratio_S")),
                      variant = factor(variant, levels=variants$v))

(plot_fig6 <- dr6 %>% ggplot(aes(x=timepoint, y=ratio, fill=subpopulation)) +
  geom_area(position="stack") + facet_grid(replicate~variant) +
  labs(y="proportion of total population") +
  scale_fill_manual(values=c("grey90","grey60","grey30","black"),
                    labels=c("recipient","transconjugant","donor","segregant"),
                    name="") +
  theme(legend.position="bottom"))
```

Plot also total counts, using a log scale.

```{r}
(plot_fig6b <- bind_rows(d_anc, df6) %>% mutate(total_pop = cfu.g_D + cfu.g_R) %>%
  ggplot(aes(x=timepoint, y=total_pop, group=replicate)) +
  geom_line(colour="black", size=0.2) + facet_grid(replicate~variant) +
  geom_hline(yintercept = 5e8, linetype="dotted", size=0.2) +
  scale_y_continuous(trans="log10", name=expression(paste("log"[10], " cfu/g")), limits=c(1e7,2e9)))
```

These do not vary greatly over the course of the experiment, and remain close to the grand mean (dotted line).

Output as `.svg`.

```{r}
svglite::svglite(height=3.5, width=7.2, file = "../plots/Fig6.svg")
plot_fig6 + theme_pub() + theme(legend.position="bottom")
dev.off()

svglite::svglite(height=3.5, width=7.2, file = "../plots/Fig6b.svg")
plot_fig6b + theme_pub()
dev.off()
```


#### Analysis

Analyse data by looking at cumulative plasmid presence over the course of the experiment. 

Calculate this from the ratios. The fact that I am looking at the ratios doesn't matter greatly, as the overall counts are similar across the experiment, and ratios directly reflect what is being presented in the figure, i.e. I am comparing the dark grey areas.

```{r}
dr6_summ <- dr6 %>% filter(subpopulation %in% c("ratio_D","ratio_T")) %>%
  arrange(replicate, timepoint, variant, subpopulation) %>%
  mutate(prop_plasmid = ratio + lag(ratio),
         v = sub(" ", "", variant)) %>%
  filter(subpopulation == "ratio_D") %>%
  group_by(replicate, v, marker) %>%
  summarise(cumulative_plasmid = sum(prop_plasmid)) %>%
  left_join(variants, by="v")
  
```

Plot this data to see if it makes sense.

```{r}
ggplot(data=dr6_summ, aes(x=v, y=cumulative_plasmid)) + geom_point() +
  facet_wrap(~PQ5759, scales="free_x")
```

Looks fine, test with a model.

```{r}
library(nlme)

mod6_1 <- gls(data=dr6_summ, cumulative_plasmid ~ PQ5759, method="REML")
mod6_2 <- lme(data=dr6_summ, cumulative_plasmid ~ PQ5759, 
              random = ~1|v, method="REML")

anova(mod6_1, mod6_2)
```

Strong random effect.

```{r}
data.frame(intercept = ranef(mod6_2), 
           v = rownames(ranef(mod6_2))) %>% 
  tibble %>%
  arrange(-abs(X.Intercept.)) %>% select(v, X.Intercept.) %>% kable()
```

Driven (as expected) by C01.

Generate plots for model validation.

```{r}
dr6_summ <- dr6_summ %>% 
  ungroup() %>% 
  mutate(mod6_2_resid  = resid(mod6_2, type = "normalized"),
         mod6_2_fitted = fitted(mod6_2))

ggplot(data=dr6_summ, aes(x=mod6_2_fitted, y=mod6_2_resid)) + 
  geom_point() + labs(x="fitted values", y="residuals")

ggplot(data=dr6_summ, aes(x=PQ5759, y=mod6_2_resid)) + 
  geom_boxplot() + labs(x="PQ5759 variant", y="residuals")

ggplot(data=dr6_summ, aes(sample=mod6_2_resid)) + stat_qq() + stat_qq_line()
```

These look fine.

The QQ plot looks like it might be significant.

```{r}
shapiro.test(resid(mod6_2, type="normalized"))
```

It is. Transform with Box-Cox and repeat analysis.

```{r}
MASS::boxcox(cumulative_plasmid ~ PQ5759, data=dr6_summ)
```

Lambda at 0.5.

```{r}
lambda <- 0.5

dr6_summ <- dr6_summ %>%
  mutate(cp_boxcox = ((cumulative_plasmid^lambda) - 1) / lambda)

mod6_3 <- gls(data=dr6_summ, cp_boxcox ~ PQ5759, method="REML")
mod6_4 <- lme(data=dr6_summ, cp_boxcox ~ PQ5759, 
              random = ~1|v, method="REML")

anova(mod6_3, mod6_4)
```

Random effects still required.

```{r}
dr6_summ <- dr6_summ %>% 
  mutate(mod6_4_resid  = resid(mod6_4, type = "normalized"),
         mod6_4_fitted = fitted(mod6_4))

ggplot(data=dr6_summ, aes(x=mod6_4_fitted, y=mod6_4_resid)) + 
  geom_point() + labs(x="fitted values", y="residuals")

ggplot(data=dr6_summ, aes(x=PQ5759, y=mod6_4_resid)) + 
  geom_boxplot() + labs(x="PQ5759 variant", y="residuals")

ggplot(data=dr6_summ, aes(sample=mod6_4_resid)) + stat_qq() + stat_qq_line()

shapiro.test(resid(mod6_4))
```

Good, proceed to model reduction.

```{r}
mod6_4_ml <- update(mod6_4, method="ML")

mod6_5_ml <- update(mod6_4_ml, .~.-PQ5759)

anova(mod6_4_ml, mod6_5_ml)
```

Shows that there is a significant effect of PQBR57_0059 variant.

```{r}
summary(mod6_4)
```

A mutated PQBR57_0059 is associated with an increase in cumulative plasmid. 

#### Tests

Do the results hold if the ancestral strain is not included?

```{r}
drop1(lme(cp_boxcox ~ PQ5759, random = ~1|v, method="ML", data=filter(dr6_summ, v != "unevolved")),
      test="Chisq")
```

Yes, though it is approaching the margin!

---

**[Back to index.](COMPMUT_index.md)**
